Copy number inference with InferCNV

### 2024-05-20 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
scale proportion by number of genes

# png ref files 
png_files <- file.path(all_results_dir_list, glue::glue("{params$library_id}_infercnv.png")) |> 
  purrr::set_names(names(all_results_dir_list))

# with ref cnv metadata file 
cnv_metadata_file <- file.path(results_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-metadata.tsv"))
```

@@

# read in cnv metadata 
cnv_metadata_df <- readr::read_tsv(cnv_metadata_file)

# read in tumor normal classifications 
manual_classifications_df <- readr::read_tsv(classifications_file)
```

@@

@@

# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL000219.1"))
cnv_df <- coldata_df[, chr_columns_to_keep]
```

@@

One approach to classifying tumor cells is to look for cells that have a lot of CNVs. 
Let's start by just summing up the total CNVs/cell and plotting that.

```{r}
# grab any columns that have has_cnv 
has_cnv_cols <- names(cnv_df)[startsWith(colnames(cnv_df), "has_cnv")]

cnv_df <- cnv_df |>
  # get total number of chromosomes with cnvs per cell and then scale that number 
  dplyr::mutate(has_cnv = rowSums(across(has_cnv_cols)),
                scaled_cnv = scale(has_cnv))
```

```{r}
# plot the total number of CNVs
ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = has_cnv)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  scale_color_viridis_c()
```
This shows that the total CNVs are highest in the group of cells that correspond to the same cells that are manually annotated as tumor cells, which is promising. 
If we were to use total CNVs we would need some sort of cut off for the number. 
Let's look at the distribution of total CNVs and also the scaled total CNVs to see if we can identify a cut off.

@@

ggplot(cnv_df, aes(x= scaled_cnv)) + 
  geom_density() +
  theme_bw() +
  labs(x = "z-scaled chromosome CNV count")
```

@@

```{r}
# add cnv classification 
cnv_df <- cnv_df |> 
  dplyr::mutate(cnv_classification = dplyr::if_else(scaled_cnv > 0, "Tumor", "Normal"))

@@

# get mean proportion of cnvs per cell and then scale that number 
cnv_df <- cnv_df |> 
  dplyr::mutate(mean_proportion = rowMeans(across(proportion_cols)),
                scaled_mean_proportion = scale(mean_proportion))
```

```{r}
ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = mean_proportion)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  scale_color_viridis_c()

ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = scaled_mean_proportion)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  scale_color_viridis_c()
```

```{r}
ggplot(cnv_df, aes(x = scaled_mean_proportion)) +
  geom_density() +
  theme_bw()
```
It looks like maybe we could again use 0 as the cutoff here and anything above 0 is a tumor cell. 
Let's see how this matches up with the tumor/normal classifications we defined just using total CNVs.

```{r}
ggplot(cnv_df, aes(x = scaled_mean_proportion, color = cnv_classification)) +
  geom_density() +
  theme_bw()
```

Interestingly, the cells we labeled as tumor have a higher mean proportion value then those that we labeled as normal. 
I'm not really sure which is "better" to use the mean proportion or just the detection of CNVs to call tumor cells.

For the rest of the notebook, we will use the classifications of tumor cells based on the sum of the total CNV occurrences.

@@

```{r}
all_classifications_df <- cnv_df |> 
  dplyr::left_join(manual_classifications_df) |>
  # relevel classifications for confusion matrix 
  dplyr::mutate(
    cnv_classification = forcats::fct_relevel(cnv_classification, "Tumor"),
    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
  )

@@

# create jaccard matrices for SingleR and CellAssign compared to aneuploid/diploid 
jaccard_matrices <- celltype_columns |>
  purrr::map(\(name) {
    make_jaccard_matrix(
      cnv_df,
      "cnv_classification",
      name
    )
  }) |> 
  purrr::set_names("SingleR", "CellAssign")
```

>>>>

# png ref files 
png_files <- all_results_dir_list |>
  purrr::map_chr(\(dir) file.path(dir, glue::glue("{params$library_id}_infercnv.png"))) |> 
  # make sure png files exists
  purrr::map(\(file)
    stopifnot("Missing PNG file" = file.exists(file))
  )

# with ref cnv metadata file 
cnv_metadata_file <- file.path(results_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-metadata.tsv"))

# full infercnv object with information on genes/ chr to use later 
final_cnv_obj_file <- file.path(results_dir, "endothelial_reference", glue::glue("{params$library_id}_cnv-obj.rds"))
```

@@

# read in cnv metadata 
# use read.delim instead of readr because we want the rownames to be read in properly 
cnv_metadata_df <- read.delim(cnv_metadata_file) |> 
  tibble::rownames_to_column("barcodes")

# read in tumor normal classifications 
manual_classifications_df <- readr::read_tsv(classifications_file)

# read in cnv object 
final_cnv_obj <- readr::read_rds(final_cnv_obj_file)
```

@@

From the [`InferCNV` documentation](https://github.com/broadinstitute/inferCNV/wiki/Interpreting-the-figure):

> The normal cells in the top heatmap define baseline expression for genes in normal cells. This baseline distribution of normal gene expression is subtracted from both the normal cells from which was defined as well as the tumor cells. Afterwards, the normal cell expression heatmap should be largely devoid of signal, with the exception of certain outlier gene expression values in certain cells. Removing this baseline of normal expression signal from the tumor cells should reveal those chromosomal regions that have significantly more or less expression than the normal cells, highlighting likely amplified or deleted whole chromosomes or large chromosomal regions.

The dendogram on the left side is hierarchical clustering of all non-reference cells. 
The color annotation next to it indicates subdivisions of the dendogram by cutting it in `k_obs_groups` groups, with the default `k_obs_groups` = 1.

@@

# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL000219.1|chrGL000194.1"))
all_cnv_df <- coldata_df[, chr_columns_to_keep]
```

@@

One approach to classifying tumor cells is to look for cells that have a lot of CNVs. 
Let's start by just summing up the total number of chromosomes with CNVs per cell and plotting that.

```{r}
# grab any columns that have has_cnv 
has_cnv_cols <- names(all_cnv_df)[startsWith(colnames(all_cnv_df), "has_cnv")]

reference_cells_only <- all_cnv_df |> 
  dplyr::filter(stringr::str_detect(subcluster, "reference")) |> 
  dplyr::mutate(has_cnv = NA,
                z_scaled_cnv = NA)

unknown_cells_only <- all_cnv_df |>
  dplyr::filter(stringr::str_detect(subcluster, "unknown")) |> 
  # get total number of chromosomes with cnvs per cell and then scale that number 
  dplyr::mutate(has_cnv = rowSums(across(has_cnv_cols)),
                z_scaled_cnv = scale(has_cnv))

# combine back together for plotting 
cnv_df <- bind_rows(reference_cells_only, unknown_cells_only)
```

```{r}
# plot the total number of CNVs
ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = has_cnv)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  scale_color_viridis_c()
```
This shows that the total CNVs are highest in the group of cells that correspond to the same cells that are manually annotated as tumor cells, which is promising. 
Those cells in grey (bottom left) are cells used as a reference. 
If we were to use total CNVs we would need some sort of cut off for the number. 
Let's look at the distribution of total CNVs and also the scaled total CNVs to see if we can identify a cut off.

@@

ggplot(cnv_df, aes(x= z_scaled_cnv)) + 
  geom_density() +
  theme_bw() +
  labs(x = "z-scaled chromosome CNV count")
```

@@

```{r}
# add cnv classification 
cnv_df <- cnv_df |> 
  dplyr::mutate(cnv_classification = dplyr::if_else(z_scaled_cnv > 0, "Tumor", "Normal"))

@@

# get total number of genes per chromosome to use in weighted mean 
chr_weights <- final_cnv_obj@gene_order |> 
  as.data.frame() |> 
  dplyr::count(chr) |> 
  # only keep chr 1-22, drops MT and GL chr
  dplyr::filter(chr %in% glue::glue("chr{seq(1,22)}")) |> 
  dplyr::pull(n)

# get mean proportion of cnvs per cell weighted by chromsome size (number of genes)
cnv_df <- cnv_df |> 
  dplyr::rowwise() |> 
  # calculate weighted mean, excluding normal cells 
  dplyr::mutate(mean_proportion = ifelse(!is.na(has_cnv), weighted.mean(across(proportion_cols), num_genes_df$n), NA)) |> 
  dplyr::ungroup()
```

```{r}
ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = mean_proportion)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  scale_color_viridis_c()
```

```{r}
ggplot(cnv_df, aes(x = mean_proportion)) +
  geom_density() +
  theme_bw()
```
It looks like we could use a cutoff of 0.1 here to classify tumor/normal cells. 
Let's see how this matches up with the tumor/normal classifications we defined just using total CNVs.

```{r}
ggplot(cnv_df, aes(x = mean_proportion, color = cnv_classification)) +
  geom_density() +
  theme_bw()
```

As expected, the cells we labeled as tumor have a higher mean proportion value then those that we labeled as normal. 
I'm not really sure which is "better" to use the mean proportion or just the detection of CNVs to call tumor cells.

```{r}
# add cnv classification 
cnv_df <- cnv_df |> 
  dplyr::mutate(cnv_mean_classification = dplyr::if_else(mean_proportion > 0.1, "Tumor", "Normal"))

ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = cnv_mean_classification)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() 
```

For the rest of the notebook, we will use the classifications of tumor cells based on the mean proportion of total CNV occurrences.

@@

```{r}
all_classifications_df <- cnv_df |> 
  dplyr::left_join(manual_classifications_df) |>
  # relevel classifications for confusion matrix 
  dplyr::mutate(
    cnv_classification = forcats::fct_relevel(cnv_mean_classification, "Tumor"),
    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
  )

@@

# create jaccard matrices for SingleR and CellAssign compared to aneuploid/diploid 
jaccard_matrices <- celltype_columns |>
  purrr::map(\(name) {
    make_jaccard_matrix(
      cnv_df,
      "cnv_mean_classification",
      name
    )
  }) |> 
  purrr::set_names("SingleR", "CellAssign")
```
